---
title: "How to downscale data from a new model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to downscale data from a new model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## starting point
You have regional (i.e non-gridded) land use data produced by an integrated assessment
model (we use "CoolNewModel" as placeholder name here) that should be harmonized and
downscaled using LUH3 as historical high resolution reference dataset.

This tutorial describes how to modify mrdownscale to process CoolNewModel data, but does
not cover harmonization or downscaling of nonland variables like fertilizer, even
though mrdownscale is capable of handling those (implemented for MAgPIE).

## requirements/preparation

### needed files
- CoolNewModel dataset file (in a format R can read)
- region mapping (which country/gridcell belongs to which region?)

### software setup
- install R: https://cran.rstudio.com/
- open R and find your home directory with
```{r, echo = TRUE, eval = FALSE}
normalizePath("~")
```
- in that folder create/edit a file called ".Rprofile" and put in the
following, so you can install mrdownscale, and madrat knows where to
store its data:
```{r, echo = TRUE, eval = FALSE}
options(repos = c(runiverse = "https://pik-piam.r-universe.dev",
                  CRAN = "https://cran.rstudio.com/"))
options(MADRAT_MAINFOLDER = file.path(normalizePath("~"), "madrat_mainfolder"))
```

- start a fresh R session and install mrdownscale incl. dependencies, pkgload for development:
```{r, echo = TRUE, eval = FALSE}
install.packages(c("mrdownscale", "pkgload"))
```

## integrating into mrdownscale
Currently mrdownscale itself needs to be modified to integrate a new model.
To do so, we recommend creating a fork of the
[mrdownscale repo](https://github.com/pik-piam/mrdownscale), then clone
your fork to a folder on your local machine. Apply the changes explained
in the following sections in that folder. To test your changes, always start
a fresh R session (!) in that folder and run `pkgload::load_all()` to
load your local mrdownscale directly from the source files.

To get an overview what changes are needed check
[this PR](https://github.com/pik-piam/mrdownscale/pull/38) which enabled
mrdownscale to process data from the COFFEE model.

### reference mapping
A mapping from CoolNewModel variables to fine reference variables
is needed in order to map from CoolNewModel variables to the variables
of any of the supported target datasets. In case CoolNewModel variables
do not cover the entire land area, e.g. if urban is missing, map these
to a variable called "rest". Put your mapping csv file into
inst/extdata/referenceMappings. Check the reference mapping for COFFEE
for an example (that file is also located in inst/extdata/referenceMappings):

```{r, echo = TRUE}
cat(readLines(system.file("extdata/referenceMappings/coffee.csv",
                          package = "mrdownscale"), n = 30), sep = "\n")
```


Some more details on how the variable mapping works (feel free to skip):
```{r, echo = TRUE}
?mrdownscale:::calcLandInputRecategorized
```

### read function
mrdownscale uses the madrat framework, if you want to learn more about madrat check this
[vignette](https://pik-piam.r-universe.dev/articles/madrat/madrat.html). Reading that
vignette should not be necessary to follow this tutorial though.

Before you can read your data via madrat your files need to be put into the
expected place:
```{r, echo = TRUE}
madrat::getConfig("sourcefolder", verbose = FALSE)
```
In this folder there's a subfolder for each dataset that can be read via
madrat. Create a new folder called CoolNewModel there and put your data
file and the region mapping file into it.


Write a function that reads in your data as a data frame. The function name
must start with `read` and the rest of the function name must match the name
of the source folder you created earlier, so call it `readCoolNewModel`.
For an example read function check readWITCH:
```{r, echo = TRUE}
mrdownscale:::readWITCH
```

To test your source folder and read function setup, start a new R session and run the following:
```{r, echo = TRUE, eval = FALSE}
pkgload::load_all("/path/to/mrdownscale")
x <- readSource("CoolNewModel", subtype = "data")
print(x)
map <- readSource("CoolNewModel", subtype = "resolutionMapping")
print(map)
```
`readCoolNewModel` is not (and should never be) called directly,
but instead indirectly via `readSource` which will, among other things,
change the working directoy to the corresponding source folder
first and also store results into the cache and load them from there if
nothing relevant changed.


### calcLandInput
ensure values >= 0, not NA/missing
ensure non-overlapping categories
ensure total area is constant over time (might need to add "rest" variable)
convert to magclass
variable names must match those in reference mapping

### calcResolutionMapping
based on your region mapping, create mapping between regional level and target grid level

### toolLandCategoriesMapping
add CoolNewModel to if-else like the other models

## running harmonization and downscaling
```{r, echo = TRUE, eval = FALSE}
calcOutput("LandHarmonized")
retrieveData("SCENARIOMIP", input = "CoolNewModel")
```
