---
title: "How to downscale data from a new model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to downscale data from a new model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## starting point
You have regional (i.e non-gridded) land use data produced by integrated assessment
model (we use "yourmodel" as placeholder name here) that should be harmonized and
downscaled. This requires a high resolution reference/target dataset, currently
LUH3, LUH2v2h, and landuseinit (MAgPIE initialization data) are available.

It is assumed that data from yourmodel was not processed in mrdownscale
before (currently MAgPIE, WITCH, COFFEE are already supported). This tutorial does
not cover harmonization or downscaling of nonland variables like fertilizer, even
though mrdownscale is capable of handling those (implemented for MAgPIE).

## requirements/preparation
- dataset file (in a format R can read)
- region mapping (from country/gridcell to region)
- install R: https://cran.rstudio.com/

```{r, echo = TRUE, eval = FALSE}
options(repos = c(runiverse = "https://pik-piam.r-universe.dev",
                  CRAN = "https://cran.rstudio.com/"))
install.packages(c("mrdownscale", "pkgload"))
```

## integrating into mrdownscale
see the following PR how this looked like for COFFEE
https://github.com/pik-piam/mrdownscale/pull/38

### reference mapping
create map from non-overlapping land variable names to fine reference variables
link to other documentation that explains how category remapping works

### read/correct/convert
reference madrat vignette
read in data

### calcLandInput
ensure values >= 0, not NA/missing
ensure non-overlapping categories
ensure total area is constant over time (might need to add "rest" variable)
convert to magclass
variable names must match those in reference mapping

### calcResolutionMapping
based on your region mapping, create mapping between regional level and target grid level

### toolLandCategoriesMapping
add yourmodel to if-else like the other models

## running harmonization and downscaling
```{r, echo = TRUE, eval = FALSE}
calcOutput("LandHarmonized")
retrieveData("SCENARIOMIP", input = "yourmodel")
```
